---
title: "Getting Started with Multilevel Models in R"
date: "2025-06-10"
description: "A walkthrough of fitting and interpreting random-effects models using lme4, with a real dataset example."
categories: [methods, R, statistics]
---

## Introduction

Multilevel models (also called mixed-effects or hierarchical models) are essential when your data have a grouped structure — students nested within schools, respondents nested within countries, observations nested within individuals over time.

In this post I walk through a minimal but complete example using the `lme4` package. We'll use the built-in `sleepstudy` dataset to keep things reproducible.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(lme4)
library(ggplot2)
library(dplyr)

# The sleepstudy dataset:
# Reaction times (ms) of 18 subjects over 10 days of sleep deprivation
head(sleepstudy)
```

## Exploring the Data

Before fitting any model, always look at the data. Here we plot each subject's trajectory:

```{r}
#| label: fig-spaghetti
#| fig-cap: "Reaction time by days of sleep deprivation, per subject."
#| fig-height: 5

ggplot(sleepstudy, aes(x = Days, y = Reaction, group = Subject)) +
  geom_line(alpha = 0.4, color = "#1a3a5c") +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE,
              color = "#c0392b", linewidth = 1.2) +
  labs(
    x = "Days of sleep deprivation",
    y = "Reaction time (ms)",
    title = "Reaction times increase with sleep deprivation",
    subtitle = "Each line = one subject; red line = population trend"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

Two things are immediately visible: (1) there is a clear positive trend on average, and (2) subjects vary substantially in both their baseline and their rate of change.

## Fitting the Models

### Baseline: fixed effects only (ignores grouping)

```{r}
#| label: model-ols

m0 <- lm(Reaction ~ Days, data = sleepstudy)
summary(m0)
```

This model treats all 180 observations as independent — which they are not, because we have 10 observations per person.

### Model 1: random intercepts

Allow each subject to have their own baseline reaction time:

```{r}
#| label: model-ri

m1 <- lmer(Reaction ~ Days + (1 | Subject), data = sleepstudy)
summary(m1)
```

### Model 2: random intercepts + random slopes

Allow each subject to have their own baseline *and* their own rate of change:

```{r}
#| label: model-ris

m2 <- lmer(Reaction ~ Days + (1 + Days | Subject), data = sleepstudy)
summary(m2)
```

## Comparing Models

```{r}
#| label: model-comparison

anova(m1, m2)
```

The likelihood ratio test strongly favors `m2` — the random slopes model. This makes substantive sense: people differ not just in how tired they start, but in how quickly sleep deprivation affects them.

## Visualizing the Random Effects

```{r}
#| label: fig-ranef
#| fig-cap: "Subject-specific intercepts and slopes (BLUPs) from Model 2."
#| fig-height: 4

ranef_df <- as.data.frame(ranef(m2)$Subject)
ranef_df$Subject <- rownames(ranef_df)
names(ranef_df)[1:2] <- c("Intercept", "Slope")

ggplot(ranef_df, aes(x = Intercept, y = Slope, label = Subject)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(color = "#1a3a5c", size = 3) +
  ggrepel::geom_text_repel(size = 3, color = "#6b6b6b") +
  labs(
    x = "Random intercept",
    y = "Random slope (Days)",
    title = "Individual deviations from the population trend"
  ) +
  theme_minimal(base_size = 12)
```

::: {.callout-tip}
**Interpretation tip:** Subjects in the upper-right quadrant are slow *and* get slower quickly. Subjects in the lower-left are fast *and* relatively robust to sleep deprivation.
:::

## Key Takeaways

1. Ignoring the grouped structure inflates Type I error and underestimates standard errors.
2. Always plot your data before modeling — the spaghetti plot told us we needed random slopes.
3. `lmer()` syntax: `(1 | group)` = random intercept; `(1 + x | group)` = random intercept and slope.

## Going Further

- Bates et al. (2015) is the canonical reference for `lme4`.
- For Bayesian multilevel models, see `brms` — same formula syntax, much richer output.
- For a textbook treatment: Gelman & Hill (2007) *Data Analysis Using Regression and Multilevel/Hierarchical Models*.
